---
title: "MET581-Lecture-02-Explore-R-homework-answers"
author:
  - name: "Gabriel Mateus Bernardo Harrington"
  - name: "Matthew Bracher-Smith"
date: "2024-10-15"
format: html
editor: source
---

# Exploring Data in R - Answers

## Overview

Quarto is a document format that allows you to create and render documents in multiple formats including HTML, PDF, and MS Word. For more details on using Quarto, see [the Quarto website](https://quarto.org). When you render this document, it will generate output that includes both the content and the results of any embedded R code chunks. You can embed an R code chunk like this:

```{r}
#| eval: false
summary(cars)
```

This will display a summary of the cars dataset like shown below:

```{r}
#| echo: false
summary(cars)
```

For your homework, you will answer the questions below using Quarto. Make sure to include both the code and results in your knitted PDF or html file.

## Questions

1.  Use Quarto to write a paragraph explaining the tidyverse concept. Your paragraph should include:

2. Create a tibble with columns labeled: name, age, height, gender. Your tibble should contain at least 13 rows of data.

```{r}
#| echo: true
#| warning: false
library(tibble)
library(dplyr)

tib <- tibble(
  "name" = c(
    "Luke",
    "Julia",
    "Matt",
    "Simon",
    "Laura",
    "Paul",
    "Frank",
    "Chris",
    "Pam",
    "Sue",
    "Sim",
    "Tom",
    "Ben"
  ),
  "age" = c(22, 21, 18, 19, 20, 16, 17, 26, 23, 25, 28, 18, 19),
  "height" = c(182, 180, 178, 183, 180, 179, 176, 175, 180, 181, 182, 170, 175),
  "gender" = c("M", "F", "M", "M", "F", "M", "M", "M", "F", "F", "M", "M", "M")
)

tib
```

   -  2a: Use tibble code to select the data for row 4. 
   
```{r}
slice(tib, 4)
```

Or

```{r}
tib[4, ]
```

   -  2b: Retrieve the height of a person based on their name. 

```{r}
tib[tib$name == "Luke", ]$height
```

Or

```{r}
filter(tib, name == "Luke")$height
```

   -  2c: Calculate the ratio of men to women. 

```{r}
nrow(tib[tib$gender == "M", ]) / nrow(tib[tib$gender == "F", ])
```

   -  2d: Calculate the average age and the standard deviation. 
   
```{r}
mean(tib[['age']])
sd(tib[['age']])
```
  
3. Turn the dataset iris into a tibble.

```{r}
iris_tib <- as_tibble(iris)
iris_tib
```

4. Why might tibbles be preferable to traditional R data frames for data scientists?

5. List some of the most important options provided by the readr package.

6. Import the pheno_unclean.txt file using readr.

```{r}
library(readr)
pheno <- read_tsv(here::here("data/pheno.txt"))
pheno
```

7. Identify any errors in the pheno_unclean.txt file using R code.

```{r}
pheno[pheno$Phenotype > 1, ]
```

Or

```{r}
filter(pheno, Phenotype > 1)
```
