---
title: "MET581-Lecture-02-Expore-R-homework"
author: "Matthew Bracher-Smith"
date: "2024-10-15"
format: html
editor: source
---

# Exploring Data in R - Exercises

## Overview

Quarto is a document format that allows you to create and render documents in multiple formats including HTML, PDF, and MS Word. For more details on using Quarto, see [the Quarto website](https://quarto.org). When you render this document, it will generate output that includes both the content and the results of any embedded R code chunks. You can embed an R code chunk like this:

```{r}
#| eval: false
summary(cars)
```

This will display a summary of the cars dataset like shown below:

```{r}
#| echo: false
summary(cars)
```

For your homework, you will answer the questions below using Quarto. Make sure to include both the code and results in your knitted PDF or html file.

## Questions

1.  Use Quarto to write a paragraph explaining the tidyverse concept. Your paragraph should include:

   - Example code
   - Inline code
   - Blockquotes
   - Lists
   - Bold or italic text

2. Create a tibble with columns labeled: name, age, height, gender. Your tibble should contain at least 13 rows of data.

   -  2a: Use tibble code to select the data for row 4. 
   -  2b: Retrieve the height of a person based on their name. 
   -  2c: Calculate the ratio of men to women. 
   -  2d: Calculate the average age and the standard deviation. 
  
3. Turn the dataset iris into a tibble.

4. Why might tibbles be preferable to traditional R data frames for data scientists?

5. List some of the most important options provided by the readr package.

6. Import the pheno_unclean.txt file using readr.

7. Identify any errors in the pheno_unclean.txt file using R code.
