---
title: "Factors with forcats"
subtitle: "Reference material — not covered in the taught sessions"
author: "Gabriel Mateus Bernardo Harrington"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: false
execute:
  echo: true
  warning: false
---

::: {.callout-note appearance="simple"}
This page is **reference material**, not a taught session.

One part of it is not optional, though: [coercing factors](#gotchas-coercing-factors)
at the bottom. That is a bug that silently corrupts your data, has reached
publication more than once, and is one of the examples in the [AI assistants
session](../../00_llm_use/using_ai_assistants.qmd) precisely because it does not
error. Read that section even if you skip the rest.
:::

## Setup

```{r}
#| output: false
library(forcats)
library(ggplot2)
library(dplyr)
library(gapminder)
```

`gss_cat` used throughout is a sample of the US General Social Survey that ships
with `forcats`; `diamonds` ships with `ggplot2`.

## What is a factor?

A factor:

- is how we store categorical variables in R
- contains a fixed and known set of possible values, called its **levels**

Create one with `factor()`, which takes `factor(vector, levels, labels)`:

```{r}
factor(c(0, 1, 1, 1, 0), labels = c('Female', 'Male'))
```

## Making factors

Factors can carry an inherent order — but the default may not be the order you
want.

```{r}
data <- factor(c("Dec", "Jun", "Apr"))
data
```

Sorting may not give what you expect:

```{r}
sort(data)
```

That is because:

- factors always have an *internal* order, even if you don't specify one
- if you don't set the levels, they are **alphabetical**
- if you want a specific order, you have to say so

```{r}
monthLevels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                 "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
data <- factor(c("Dec", "Jun", "Apr"), levels = monthLevels)
sort(data)
```

### Values outside your levels disappear silently

```{r}
factor(c("Dec", "Jum", "Apr"), levels = monthLevels)
```

Note the `NA`. A typo in your data becomes a missing value with **no warning at
all**. By contrast, `readr::parse_factor()` tells you:

```{r}
readr::parse_factor(c("Dec", "Jum", "Apr"), levels = monthLevels)
```

::: {.callout-important appearance="simple"}
This is the pattern to internalise: the base R route fails silently, the tidyverse
route complains. When you are reading in real categorical data, prefer the one
that complains.
:::

## Why bother with forcats?

- It covers the common things you actually need to do with factors.
- It works well with `ggplot2` — same authors.
- It generally tries to warn you when something looks wrong.
- It is an anagram of "factors", and it is **for cat**egorical**s**.

### `fct_inorder()`

Sets the levels to the order they appear in the vector.

```{r}
head(fct_inorder(gss_cat$marital))
```

### `fct_relevel()`

Moves one or more levels to the front.

```{r}
head(fct_relevel(gss_cat$marital, 'Married'))
```

### `fct_recode()`

Renames existing levels by hand. Note that mapping two old levels to the same new
name merges them.

```{r}
myFactor <- factor(c("M", "F", "O", "M", "P", "M",
                     "F", "F", "F", "M", "O", "P"))
fct_recode(myFactor,
           male = "M", female = "F",
           unknown = "O", unknown = "P")
```

### `fct_reorder()`

The most useful one. It reorders factor levels by another variable, which brings
structure to a plot. Compare these two.

Without:

```{r}
gss_cat |>
  group_by(marital) |>
  summarise(tvhours = mean(tvhours, na.rm = TRUE)) |>
  ggplot(aes(tvhours, marital)) +
  geom_point()
```

With:

```{r}
gss_cat |>
  group_by(marital) |>
  summarise(tvhours = mean(tvhours, na.rm = TRUE)) |>
  ggplot(aes(tvhours, fct_reorder(marital, tvhours))) +
  geom_point()
```

### `fct_reorder2()`

Probably the second most useful. It orders the levels to match the values at the
far right of the plot, so the legend order matches the line order — which is
surprisingly helpful when reading a graph.

Without:

```{r}
gss_cat |>
  filter(!is.na(age)) |>
  count(age, marital) |>
  group_by(age) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(age, prop, colour = marital)) +
  geom_line(na.rm = TRUE)
```

With:

```{r}
gss_cat |>
  filter(!is.na(age)) |>
  count(age, marital) |>
  group_by(age) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(age, prop, colour = fct_reorder2(marital, age, prop))) +
  geom_line() +
  labs(colour = "marital")
```

### Others worth knowing

- `fct_rev()` reverses the level order
- `fct_infreq()` orders from most to least frequent
- `fct_lump()` combines the least common levels into "other"
- `fct_expand()` adds new levels
- `fct_relabel()` relabels levels programmatically

## Gotchas: coercing factors {#gotchas-coercing-factors}

Sometimes the damage is obvious:

```{r}
x <- factor(c(1, 1, 0, 0, 1, 0, 1, 1, 1, 0))
as.numeric(x)
```

Sometimes it is not:

```{r}
x <- factor(c(1, 1, 2, 5, 3, 3, 1, 6, 5, 1, 6, 2))
as.numeric(x)
```

Those are R's **internal level codes**, not your data. The values look plausible,
the vector is the right length, and nothing warns you.

::: {.callout-warning appearance="simple"}
## The rule
**Never convert a factor straight to numeric.** Convert to character first, then to
numeric:

```{r}
x <- factor(c(1, 1, 2, 5, 3, 3, 1, 6, 5, 1, 6, 2))
as.numeric(as.character(x))
```
:::

## Practice

**Basics**

- Create a factor called `marauders` containing `'moony'`, `'wormtail'`,
  `'padfoot'` and `'prongs'` in alphabetical order.
- Create a factor called `patronus` containing `'stag'`, `'dog'`, `'otter'`, with
  levels in the order they appear in the input.
- Print only the levels of each.

**Using `diamonds`**

- Use `dplyr::count()` to show how many rows there are for each level of `cut`. Is
  it any different from `forcats::fct_count()`?
- Use `dplyr::arrange(desc())` on `cut`. What did it actually sort by?

**Using `gss_cat`**

- How many levels are there in `relig`?
- Reorder the levels of `denom` by order of appearance.
- Take the code below, which plots income by age. Reorder the levels of `rincome`
  by `age`. Then instead try just moving `n/a` to the front. Which reads better?

```{r}
#| output: false
gss_cat |>
  group_by(rincome) |>
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()) |>
  ggplot(aes(age, rincome)) +
  geom_point()
```

- Rename the `partyid` levels `"Not str republican"`, `"Ind,near dem"` and
  `"Ind,near rep"` to `"Not strong republican"`, `"Independent, near democrat"`
  and `"Independent, near republican"`.
- Edit the code below so the bars are sorted lowest to highest, using
  `fct_infreq()` and `fct_rev()`.

```{r}
#| output: false
gss_cat |>
  ggplot(aes(marital)) +
  geom_bar()
```

- Change the code below so the legend colours match the order of the lines at the
  right-hand side of the plot.

```{r}
#| output: false
diamonds |>
  filter(color == 'J', depth > 55, carat <= 2.5) |>
  ggplot(aes(carat, price, col = cut)) +
  geom_line(alpha = 0.6)
```

## Further reading

- [R for Data Science 2e, chapter 17 — Factors](https://r4ds.hadley.nz/factors)
- The [`forcats` documentation](https://forcats.tidyverse.org/)
