---
title: "Strings and regular expressions"
subtitle: "Reference material — not covered in the taught sessions"
author: "Gabriel Mateus Bernardo Harrington"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: false
execute:
  echo: true
  warning: false
---

::: {.callout-note appearance="simple"}
This page is **reference material**. It is not taught in a session and you will not
be assessed on the details of regular expressions.

It is here because real data is full of messy text — inconsistent column names,
identifiers with mixed separators, categories typed five different ways — and
`stringr` is how you deal with that in the tidyverse. Come back to it when you hit
that problem in the report.
:::

## Setup

```{r}
#| output: false
library(stringr)
library(dplyr)
library(tidyr)
library(gapminder)
library(nycflights13)
```

`words` and `sentences` used below are small example datasets that ship with
`stringr`, so they are available as soon as it is loaded.

## What is a string?

- Strings are a sequence of characters, which has to be represented in memory in
  binary.
- This was first widely done using ASCII (American Standard Code for Information
  Interchange).
- But ASCII only allows for 128 characters, so nowhere near enough for all
  languages.
- We now represent strings using Unicode (which didn't appear in its current form
  until the 90s), which allows for a lot — more than a million — characters.
- UTF-8 is the most common encoding for Unicode. It uses a variable number of
  8-bit units to represent characters.
- R uses UTF-8 by default (though it will report the label "unknown" unless
  non-ASCII characters appear), and `stringr` fully supports Unicode.

## stringr

- Removes inconsistencies found in base R.
- Built on top of the `stringi` package.
- Starts all functions with `str_`, so autocomplete is genuinely useful.
- More advanced use requires regular expressions.

## Regular expressions

As quoted by [R for Data Science, 1st
edition](http://r4ds.had.co.nz/strings.html):

> *"When you first look at a regexp, you'll think a cat walked across your
> keyboard, but as your understanding improves they will soon start to make
> sense."*

If you've read about regex before, you will also have come across:

> *"Some people, when confronted with a problem, think 'I know, I'll use regular
> expressions.' Now they have two problems."*

### The short version

- They allow you to match patterns in strings.
- The most basic form matches an actual chunk of text, e.g. `hag` in `hagrid`.
- Match classes of character with `\s`, `\d`, `\w`, `[abc]` and `[^abc]`.
- **Because we're in R, you have to write `\\` instead of `\`** — so `\\w`, `\\d`.
  This catches everybody at least once.
- Match any character at all with `.`.
- Expand a match with `?`, `+`, `*` or `{n,m}`.
- Anchor it to the start with `^` or the end with `$`.
- So `^\\w+_\\d{4}$` matches `"hagrid_2020"` but not `"hagrid_120"`.

The rest of this section works through each of those on one example vector.

### Matching literal text

```{r}
my_string <- c(
  'Hagrid',
  'Hermione',
  'Harry.Potter',
  'Ronald_Weasley',
  '24xHouse Elves'
)
str_detect(my_string, 'Hag')
```

### Character classes

Match types of character with `\\w` (word), `\\d` (digit) and `\\s` (whitespace).

```{r}
my_string
str_detect(my_string, '\\w')
```

```{r}
str_detect(my_string, '\\d')
```

```{r}
str_detect(my_string, '\\s')
```

### Matching any character

`.` matches anything, which is why it matches every element here.

```{r}
str_detect(my_string, '.')
```

### Quantifiers

`?`, `+`, `*` and `{n,m}` control how many times the preceding pattern may repeat.

```{r}
str_detect(my_string, '\\w+')
```

Note the difference `[xyz]` versus `[^xyz]` makes — the second negates the class.

```{r}
str_detect(my_string, '\\d{2}[xyz]\\w+\\s?\\w+')
str_detect(my_string, '\\d{2}[^xyz]\\w+\\s?\\w+')
```

### Anchors

`^` anchors to the start of the string, `$` to the end. Anchoring is usually what
turns a pattern that *nearly* works into one that does.

```{r}
str_detect(my_string, '^H\\w+$')
```

```{r}
str_detect(my_string, '^Hag$')
```

## Basic stringr operations

### String length and counts

```{r}
# number of characters in each string
str_length(my_string)

# number of matches within each string
str_count(my_string, 'Ha')
```

### Concatenating strings

```{r}
str_c('this', 'that', sep = ', ')
```

## Manipulating strings

### Extracting and replacing

```{r}
# extract substrings by position
str_sub(my_string, 1, 3)
```

```{r}
# replace the first match in each string
str_replace(my_string, '[\\._x]', ' ')
```

::: {.callout-tip appearance="simple"}
`str_replace()` replaces only the **first** match in each string.
`str_replace_all()` replaces every match. Picking the wrong one is a quiet bug of
exactly the kind this module keeps warning you about — it will not error.
:::

### Detecting and viewing

```{r}
# returns a logical vector - the tidyverse alternative to grepl()
str_detect(my_string, 'Hagrid')
```

```{r}
# shows you what your pattern actually matched
str_view(my_string, '^\\w')
```

`str_view()` is the single most useful debugging tool here. When a pattern isn't
working, look at what it *is* matching before changing it.

### Splitting and sorting

```{r}
sentences |>
  head(1) |>
  str_split(" ")
```

```{r}
str_sort(words[1:10], locale = 'en')
```

Always set `locale` when sorting. Sort order is language-dependent, and leaving it
implicit makes your results depend on the machine they ran on — which is exactly
what reproducibility is meant to rule out.

### Changing case

```{r}
str_to_upper(words[1:10])
str_to_lower(words[1:10])
str_to_sentence(words[1:10])
str_to_title(words[1:10])
```

```{r}
sentences |>
  head(1) |>
  str_to_sentence()

sentences |>
  head(1) |>
  str_to_title()
```

## Practice

If you want to work through this material rather than just read it:

**Concatenating and replacing**

- Concatenate the strings `"day to"` and `"day"`, separated by a hyphen.
- Using `starwars`: select the `hair_color` column and replace `', '` with `'/'`.
  Hint: pipe a column into `pull()` to get a vector for `stringr` to work on.
- Using `flights`: select the columns ending in `delay` and remove the underscore
  from those column names.

**Splitting**

- Split `"Harry, did you put your name in the Goblet of Fire?"` into its
  components.
- Use `boundary("word")` instead of `" "` and compare the results.
- Using the fifth line of `sentences`: split by word boundary, convert to
  lowercase and sort. Hint: `unlist()`.

**Case**

- Using `starwars`: convert `hair_color` values to Sentence case, and all column
  names to Title Case.
- Using `gapminder`: remove any camelCase from the column names (all to lower
  case), and convert the `continent` values to upper case.

## Further reading

- [R for Data Science 2e](https://r4ds.hadley.nz/), chapters 14 and 15
- The [`stringr` documentation](https://stringr.tidyverse.org/), and
  `vignette("regular-expressions", package = "stringr")`
- [This blog
  post](https://www.joelonsoftware.com/2003/10/08/the-absolute-minimum-every-software-developer-absolutely-positively-must-know-about-unicode-and-character-sets-no-excuses/)
  by Joel Spolsky on Unicode and character sets
- [regex101.com](https://regex101.com/) — build and test a pattern interactively,
  with an explanation of each part. Remember to double your backslashes when you
  bring the pattern back into R.
