---
title: "Data Visualisation in R"
author:
  - name: "Gabriel Mateus Bernardo Harrington"
    orcid: 0000-0001-6075-3619
    email: bernardo-harringtong@cardiff.ac.uk
    affiliations: 
      - ref: cardiff
      - ref: dri
    degrees: 
      - PhD
affiliations:
  - id: cardiff
    name: Cardiff University
    city: Cardiff
    url: www.cardiff.ac.uk
  - id: dri
    name: UK Dementia Research Institue
    url: https://ukdri.ac.uk/
format: 
  revealjs:
    logo: /images/combined_logos.png
    footer: "MET581 - Data Visualisation in R"
    theme: [dark, ../styles.scss]
    include-in-header: 
      text: |
        <style>
        .center-xy {
          margin: 0;
          position: absolute;
          top: 40%;
          left: 40%;
          -ms-transform: translateY(-50%), translateX(-50%);
          transform: translateY(-50%), translateX(-50%);
        }
        </style>
execute: 
  echo: true
---

##  {.center background-color="black" background-image="/images/02_visualisation.png" background-size="contain"}

::: center-xy
### [Visualisation]{style="background-color:#000000;"}

[Making you data come to life!]{style="background-color:#000000;"}
:::

## Objectives

-   How data can be visualised in R
-   Learn how to adjust graphics
-   Get familiar with `ggplot2` and it's grammar of graphics to create amazing figures

## Data Visualisation - An Overview

-   The overall aim of visualising data:
    -   Make all your plots as self explanatory as possible!
-   For these lectures, we will focus on `ggplot2`, a tidyverse package
    -   Inspired on the [Grammer of Graphics](https://link.springer.com/book/10.1007/0-387-28695-0), a book that aims to formalise visualisations into layers
-   Has multiple add-ons, such as `ggrepel` (for text labels) and `ggpubr` (for publication-ready plots) or `patchwork` (for combining multiple plots)

## Base R plots

-   Graphs can be easily generated with the base R syntax

```{r}
#| output-location: column
#| fig-show: hold
data <- c(2,3,6,4,9)
plot(data)

plot(data, type = "l")
```

## Base R plots

-   Graphs can be easily generated with the base R syntax

```{r}
#| output-location: column
#| fig-show: hold
plot(data, type = "o", col = "blue")

barplot(data)
```

## ggplot2

-   `ggplot2` is a lot more useful and user friendly than base R, making plots look a lot nicer and with more options for building and displaying graphics
-   We'll start with an old favourite, the `mtcars` dataset!

```{r}
#| output-location: column
# Load packages
# ggplot2 is in the tidyverse
library(tidyverse) 
library(conflicted)
library(ggrepel)
library(ggstatsplot)
library(plotly)
conflicts_prefer(dplyr::filter)

mtcars |>
  ggplot(aes(x = mpg, y = hp)) +
  geom_point()
```

## ggplot2 - layers

-   `ggplot()` is the main function, and this creates the initial ggplot object where we then add multiple **layers**
-   A layer is a collection of geometric elements (**geoms**) and statistical transformations
-   An easy example of a "geom" element layer is `geom_point()`, which adds a scatter plot
-   Aesthetic mappings (**aes**) are specified with `aes()`. This is how variables (columns) in the input data are mapped to visual, or "aesthetic", properties.
-   You can give **global** "aesthetics" to a plot (will appear in every layer) by specifying this in the `ggplot()` function, or **local** aesthetics in individual layers (such as in geom_point()).

## Themes

-   A "theme" controls the finer points of the plot, like the font size and background colour
-   This is essentially customising the non-data elements
-   For example, change the default grey background to white background[^1]

[^1]: the default is `theme_grey()`

```{r}
#| output-location: column
mtcars |>
  ggplot(aes(x = mpg, y = hp)) +
  geom_point() +
  theme_bw()
```

## Themes - Global

-   These themes can be set globally, so for all plots, using the `theme_set()` function

```{r}
#| output-location: column
theme_set(theme_bw())

mtcars |>
  ggplot(aes(x = mpg, y = hp)) +
  geom_point()
```

-   You can see some more themes provided by ggplot2 [here](https://r4ds.had.co.nz/graphics-for-communication.html#themes)

## Quick plots with qplot

-   A quicker version of ggplot! Good for very basic figures

```{r}
#| output-location: column
qplot(mpg, hp, data = mtcars)
```

## Aesthetics: Colour

-   We can visualise more information by colouring the data points by another variable
-   For example, in mtcars we can map the number of cylinders to the `colour` aesthetic (or `color` if you want to spell it wrong...)

```{r}
#| output-location: column
ggplot(data = mtcars,
       mapping = aes(x = mpg, y = hp, 
                     colour = factor(cyl))) +
  geom_point()
```

## Aesthetics: Size and Shape

-   We can also map the number of cylinders to the `size` or `shape` aesthetic

```{r}
#| output-location: column
#| fig-show: hold
# Using size
ggplot(data = mtcars,
       mapping = aes(x = mpg, y = hp, 
                     size = cyl)) +
  geom_point()

# Using shape
ggplot(data = mtcars, 
       mapping = aes(x = mpg, y = hp, 
                     shape = factor(cyl))) +
  geom_point()
```

## Aesthetics: Shape & Colour

-   We can combine aesthetics as we like too!

```{r}
#| output-location: column
ggplot(data = mtcars, 
       mapping = aes(x = mpg, y = hp, 
                     shape = factor(cyl),
                     colour = factor(cyl))) +
  geom_point()
```

## Aesthetics: Conditional Colour

-   We can even map an aesthetic to a datapoint based on a **condition**, i.e. only change colour when a certain condition is met.
-   For example here, the colour varies depending on whether the car has 4 cylinders or not (`cyl == 4` being `TRUE` or `FALSE`)

```{r}
#| output-location: column
ggplot(data = mtcars,
       mapping = aes(x = mpg, y = hp,
                     colour = cyl == 4)) +
    geom_point()
```

## Aesthetics: Fill

-   Fill is yet another aesthetic

```{r}
#| output-location: column
ggplot(data = diamonds) +
    geom_bar(mapping = aes(x = cut,
                           fill = cut))
```

## Facets

-   A "facet" is one section of something that has many sections.
-   A "facet" in ggplot allows you to break up the data into different subsets and plot individual panels based on it
-   Creates "subplots" or panel-like figures
-   Really useful when you've got categorical variables (such as gender)

```{r}
#| output-location: column
ggplot(data = mtcars,
       mapping = aes(x = mpg, y = hp,
                     colour = factor(cyl))) +
  geom_point() +
  facet_wrap(~ cyl)
```

## Facets: layout

-   We can change the layout of the “facets" locally:

```{r}
#| fig-show: hold
#| output-location: column
ggplot(data = mtcars,
       mapping = aes(x = mpg, y = hp,
                     colour = factor(cyl))) +
  geom_point() +
  facet_wrap(~ cyl, ncol = 1)
  # Note the following is equivalent
  # facet_wrap(~ cyl, dir = "v")
```

## Facet grids

-   We can combine 2 variables with `facet_grid()`

```{r}
#| output-location: column
ggplot(data = mtcars,
       mapping = aes(x = mpg, y = hp,
                     colour = factor(cyl))) +
  geom_point() +
  facet_grid(am ~ cyl)
```

## Additional Geoms

-   So far we've only used `geom_point()`, but there are naturally many more geoms we can use
-   `geom_smooth()` draws a smoothed line based on the trend of the provided data
-   They can be used individually, or layered on top of one another, which is the core of the grammer of graphics

```{r}
#| output-location: column
ggplot(data = mtcars, 
       mapping = aes(x = mpg, y = hp)) +
  geom_smooth()
```

## Combining geoms

-   Here we layer two geoms

```{r}
#| output-location: column
ggplot(data = mtcars, 
       mapping = aes(x = mpg, y = hp)) +
  geom_smooth() +
  geom_point()
```

-   Note that the order of geoms can matter! (though in this case it doesn't :P)

## Layering geoms and additional aesthetics

-   The order in which we give aesthetics can also matter

```{r}
#| output-location: column
#| fig-show: hold
ggplot(data = mtcars, 
    mapping = aes(x = mpg,
                  y = hp,
                  colour = factor(cyl))) +
    geom_smooth() +
    geom_point()

# A more sensible order?
ggplot(data = mtcars, 
    mapping = aes(x = mpg,
                  y = hp)) +
    geom_smooth() +
    geom_point(aes(colour = factor(cyl)))
```

## Statistical Transformations: count

-   Some plots transform your data internally and plot those new values instead of raw values
-   The `stat` argument of different plot types (geom functions) specifies the statistical transformation
-   For example, `geom_bar()` uses `stat = "count"` as it's default to create counts of the mapped variable (as in what a bar chart does):

```{r}
#| output-location: column
#| fig-show: hold
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))

# Worth noting: geom and stats are often interchangeable
ggplot(data = diamonds) +
  stat_count(mapping = aes(x = cut))
```

## Statistical Transformations: identity

-   Another commonly used stat is "identity" when plotting bars with heights based on raw values

```{r}
#| output-location: column
## example tibble
demo <- tribble(
  ~cut,            ~value, 
  "Fair",          1610, 
  "Good",          4906, 
  "Very Good",     12082, 
  "Premium",       13791,
  "Ideal",         21551 )

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, 
                         y = value), 
           stat = "identity")
```

## Break

## Positional Adjustments

-   The `position` argument can control how geoms occupy space

```{r}
#| output-location: column
#| fig-show: hold
# stacked 
ggplot(data = diamonds) +
    geom_bar(mapping = aes(x = cut, 
                           fill = clarity))
  
# side by side 
ggplot(data = diamonds) +
    geom_bar(mapping = aes(x = cut, 
                           fill = clarity),
             position = "dodge") 
# Note the default is position = "stack"
```

## Labels: Titles and Captions

-   Another editable part of a plot are the text labels, which we can add or modify using `labs()`

```{r}
#| output-location: column
ggplot(data = mtcars, 
       mapping = aes(x = mpg, y = hp)) +
  geom_point(aes(colour = factor(cyl))) +
  labs(title = "Title",
       subtitle = "Subtitle",
       caption = "Small Caption"
        )
```

## Labels: Axis and legends

```{r}
#| output-location: column
ggplot(data = mtcars, 
       mapping = aes(x = mpg, y = hp)) +
  geom_point(aes(colour = factor(cyl))) +
  labs(x = "Miles per gallon (mpg)", 
  y = "Horsepower (hp)", 
  colour = "Cylinders"
         )
```

## Labels: Axis alternative

```{r}
#| output-location: column
ggplot(data = mtcars, 
       mapping = aes(x = mpg, y = hp)) +
  geom_point(aes(colour = factor(cyl))) +
  xlab("Miles per gallon") +
  ylab("Horsepower")
```

## Annotations

-   To add annotation to data points, we can use `geom_text()` and `geom_label()`

```{r}
#| output-location: column
#| fig-show: hold
ggplot(data = mtcars, 
       mapping = aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  geom_text(aes(label = rownames(mtcars)))

ggplot(data = mtcars, 
       mapping = aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  geom_label(aes(label = rownames(mtcars)), 
             alpha = 0)
```

## Annotations: ggrepl

-   The `ggrepl` package can help make more legible labels

```{r}
#| output-location: column
#| fig-show: hold
ggplot(data = mtcars, 
       mapping = aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  ggrepel::geom_text_repel(aes(
    label = rownames(mtcars)), 
    max.overlaps = 100)
# Label only a subset
ggplot(data = mtcars,
       mapping = aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  ggrepel::geom_text_repel(aes(
    label = rownames(mtcars[1:3, ])),
    data = mtcars[1:3, ])
```

## Zooming

-   To control the plot limits, you have 3 methods:
    -   Adjusting the data that's plotted
    -   Setting xlim and ylim in `coord_cartesian()` **(do this!!)**
    -   Setting the limits in each scale
    
```{r}
#| output-location: column
# Not ideal - need strong justification!
mtcars |>
  filter(mpg >= 20, hp <= 150) |>
  ggplot(mapping = aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  geom_smooth()
```
    
## Zooming - coord_cartesian

```{r}
#| output-location: column
ggplot(data = mtcars, mapping = 
         aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  geom_smooth() +
  coord_cartesian(xlim = c(20, 35), 
                  ylim = c(0, 150))
```

This is the **RIGHT WAY**, using `coord_cartesian()`

## Zooming - lims

```{r}
#| output-location: column
ggplot(data = mtcars, mapping = 
         aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  geom_smooth() +
  lims(x = c(20, 40),
       y = c(0, 150))
```

This is also not ideal, as it removes data outside the limits!

## Scales

-   "scale" allows you control mapping things like colour, size and shape to data values
-   "scale" draws a legend or axes
-   `ggplot2` automatically adds default scales behind the scenes

```{r}
#| output-location: column
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl)))
```

## Scales - defaults

-   Is the same as:

```{r}
#| output-location: column
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_colour_discrete()
```

## Scales - axis breaks

-   The naming scheme tells you the aesthetic (`x_`, `y_`, `colour_`, etc) and the name of the scale (`continuous`, `discrete`)

```{r}
#| output-location: column
#| fig-show: hold
# Change the breaks on the axis
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  scale_y_continuous(
    breaks = seq(0, 350, by = 50)
    )
```

## Scales - axis labels

```{r}
#| output-location: column
#| fig-show: hold
#| code-line-numbers: "1-9|11-16"
# Adding text to labels
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  scale_y_continuous(
    breaks = seq(0, 350, by = 50),
    labels = paste0(
      "HP ", seq(0, 350, by = 50)
      )
    )

# No labels
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  scale_y_continuous(
    breaks = seq(0, 350, by = 50), 
    labels = NULL)
```

## Legend Layout - position

-   The legend can of course also be modified in lots of ways

```{r}
#| output-location: column
#| fig-show: hold
p <- ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl)))

# Legend at the top
p + theme(legend.position = "top")
# No legend
p + theme(legend.position = "none")
# Note - the default is "right"
```

## Legend Layout - guides

-   We can use the `guides()` function to control the legend display

```{r}
#| output-location: column
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(colour = factor(cyl)), 
             alpha = 0.5) +
  guides(colour = guide_legend(
    ncol = 2, 
    override.aes = list(size = 3, 
                        alpha = 1))
    )
```

## Controlling the Colour Scale - alt palettes

-   The default ggplot2 colours we get are a bit rubbish
-   Many pre-defined colour palettes are available to change that
-   Such as from the `RColorBrewer` package
-   Palette explainer [here](https://r-graph-gallery.com/38-rcolorbrewers-palettes.html) and interactive browser [here](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3)

```{r}
#| output-location: column
#| fig-show: hold
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  scale_colour_brewer(palette = "Set1")
```

## Controlling the Colour Scale - manually

```{r}
#| output-location: column
#| fig-show: hold
#| code-line-numbers: "1-6|8-16"
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  scale_colour_manual(values =
                        c("black", 
                          "pink", 
                          "turquoise"))

# Explicitly setting the values to colours
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  scale_colour_manual(values =
                        c(
                          `4` = "black",
                          `6` = "pink",
                          `8` = "red"
                        ))
```

## Saving plot - the ggplot2 way

-   We of course want to be able to save the beautiful plots we make! We can do this using `ggsave()`
-   If not specified, it will save the most recent plot we create to our disk
-   The format of the plot is defined in the filename extension (.pdf or .png for example)

```{r}
#| eval: false
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl)))
# Save the last printed plot
ggsave(filename = "my_plot_1.pdf")
# Save the plot to a variable first
plot <- ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl)))
ggsave(filename = "my_plot_1.png", plot)
```

## Saving plot - the base R way

-   With base R we need to set the device we want to save with first, then print the plot, and then close the device

```{r}
#| eval: false
# Save to png
png(filename = "my_plot.png", width = 500, height = 400)
# Print the plot
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl)))
# Close the device
dev.off()
```

## Compound plots

-   What if you want to put two or more plots together to save?

```{r}
#| output-location: column
plot1 <- ggplot(mtcars, 
                aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl)))
plot2 <- ggplot(mtcars, 
                aes(x = qsec, y = hp)) +
  geom_point(aes(color = factor(cyl)))
plot1 + plot2
```

-   See the [`patchwork`](https://patchwork.data-imaginist.com/articles/patchwork.html) or [`cowplot`](https://wilkelab.org/cowplot/articles/introduction.html) packages for more info

## Boxplots and Violin Plots

-   Boxplots and Violin Plots are very common within biolosciences(protein levels, patient data, SNP frequency etc.)
-   Be warned that boxplots can sometimes be misleading and so it's always good to check the raw data too! See [here](https://www.data-to-viz.com/caveat/boxplot.html) for more info
-   Have a go at creating your own boxplots and violin plots using the mtcars/diamonds/other datasets!

```{r}
#| output-location: column
mtcars$cyl <- as.factor(mtcars$cyl)
# Make base plot
p <- ggplot(mtcars, aes(cyl, mpg))

p + geom_boxplot(aes(colour = cyl))
```

## Boxplots and Violin Plots - examples

-   `geom_jitter` is like `geom_point`, but adds noise so point aren't on top of each other, handy for mapping the raw data onto other geoms!

```{r}
#| output-location: column
#| fig-show: hold
p + geom_boxplot(aes(fill = cyl), 
                 alpha = 0.3) +
  geom_jitter(size = 0.8)
# Violin
p + geom_violin(aes(fill = cyl))
```

## Boxplots and Violin Plots - combining geoms

- We can of course layer them on top too!

```{r}
#| output-location: column
# All three!
p + geom_violin(aes(fill = cyl), 
                width = 1.4) +
  geom_boxplot(width = 0.1, 
               colour = "grey", 
               alpha = 0.3) +
  geom_jitter(size = 0.8, 
              width = 0.1, 
              colour = "grey") +
  scale_fill_viridis_d() +
  theme_minimal()
```

## Boxplots and Violin Plots - numeric to categorical

```{r}
#| output-location: column
#| fig-show: hold
# discretise numeric data into categorical
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_boxplot(aes(
    group = cut_width(carat, 0.2)
    ))
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_boxplot(aes(
    group = cut_number(carat, 20)
    ))
```

## 3D Plots

-   `plotly` can be used to make 3D plots, but this is very situational and should probably only be done in cases where the plot is intended to be interactive

```{r}
#| output-location: column
#| fig-height: 5
#| fig-width: 5
mtcars$gear <- as.factor(mtcars$gear)

plot_ly(
  mtcars,
  x = ~ wt,
  y = ~ hp,
  z = ~ qsec,
  color = ~ gear
) |>
  add_markers() |>
  plotly::layout(scene = list(
    xaxis = list(title = 'Weight'),
    yaxis = list(title = 'Gross horsepower'),
    zaxis = list(title = '1/4 mile time')
  ))
```

## [`ggstatsplot`](https://indrajeetpatil.github.io/ggstatsplot/index.html)

- A handy way to quickly look at correlations!

```{r}
ggstatsplot::ggscatterstats(mtcars, x = hp, y = qsec)
```

## Links and Resources

-   Don't try to memorise everything, look stuff up!

-   ggplot2 cheat sheet: <http://rstudio.com/resources/cheatsheets>

-   A book on ggplot2 by the authors: <https://ggplot2-book.org>

-   <https://exts.ggplot2.tidyverse.org/gallery/>

-   <https://www.r-graph-gallery.com>

-   <https://github.com/jrnold/ggthemes>

## That's all folks! {.center}

-   These slides can be found on the website [here](https://h-mateus.github.io/MET581---Computing-for-Bioinformatics-and-Genetic-Epidemiology/):

![](/images/website_qrcode.svg){fig-align="center"}

