---
title: "Wrangling Data 2 Homework Answers"
# Unlinked from the sidebar and excluded from site search so students do not
# stumble onto the answers. The page is still published, so the URL can be shared
# deliberately once the homework has been discussed.
search: false
author: "Matthew Bracher-Smith"
date: "2024-10-18"
format: 
  html:
    number-sections: true
    highlight: tango
editor: source
execute:
  echo: true
---

```{r}
#| include: false
library(plyr)
library(dplyr)
library(nycflights13)
library(stringr)
library(data.table)
library(tibble)
```

This document contains all questions for lesson 4. Please create a quarto document containing all text, code and output used to answer the questions.

## 1. Explain how escapes work in R.  

Anything along the lines of: 

* you need to escape the string and the regex within R function. Students may also have written along the lines of 'You need to use `\\` for each escape because of the use of special characters (eg a backslash itself `\`)'.

## 2. List and describe the tidyverse commands used to join two datasets together.   
  
left_join : 

* All rows from dataset1
* All columns from dataset1 and dataset2
* If a row exists only in dataset1, data is retained.
* If a row exists only in dataset2, data is not retained.

right_join:

* all rows from dataset2
* all columns from dataset1 and dataset2
* If a row exists only in dataset1, data is not retained.
* If a row exists only in dataset2, data is retained.

inner_join:

* Only rows that exist in dataset1 and dataset2

full_join:
  
* All rows in dataset1 and dataset2
* Joining with missing or duplicate keys
* If data is missing from one dataset but is in the other, depending on the type of join you are performing, this can result in missing data (NA values).

semi_join(x, y):

* keeps all observations in x that have a match in y

anti_join(x, y):

* drops all observations in x that have a match in y

## 3. Load the nycflights13 package and join the datasets flights and airlines together, selecting only the columns: `year`, `month`, `day`, `hour`, `origin`, `tailnum` and `carrier` from flights. Can you do the same but using the mutate() function instead?

```{r}
# one example:
flights2 <- flights %>% 
  select(year:day, hour, origin, tailnum, carrier) %>%
  left_join(airlines, by = "carrier")

# part 2
flights3 <- flights %>%
  select(year:day, hour, origin, tailnum, carrier) %>%
  mutate(name = airlines$name[match(carrier, airlines$carrier)])
```

## 4. In the starwars dataset and in one command, add two new columns:

* one column that converts the `name` column to lower case
* one column that converts the `eye_colour` to uppercase

```{r}

# one example:
starwars_mutated <- starwars %>%
  mutate(name_lower = str_to_lower(name),
         eye_color_upper = str_to_upper(eye_color))
```

5. Download the gene annotation file from the NCBI found [here](https://cf-my.sharepoint.com/:u:/g/personal/smithmr5_cardiff_ac_uk/Ed6jtKhhKgRGq5R-IoQHTAYBLgO6FC_aJ6u7mydY1mpc_A?e=siJJ4A)

**unzip the file and load in `NCBI37.3.gene.loc` into R**

* add the column headers: `Entrez_Gene_ID`, `CHR`, `BP_START`, `BP_END`, `STRAND`, `GENE_NAME`
* How many Genes are on the positive strand?
* How many Genes begin and end with a letter?

```{r}
#| eval: false
# using readr to read the file
# note that we haven't correctly handled chromosomes, which needs to be specified in col_types
test <- readr::read_tsv(
  "./data/NCBI37.3/NCBI37.3.gene.loc",
  col_names = c("Entrez_Gene_ID", "CHR", "BP_START" ,"BP_END", "STRAND", "GENE_NAME"))

# using data.table to read the file
test <- fread("./data/NCBI37.3/NCBI37.3.gene.loc",header= F)
setnames(test,c("Entrez_Gene_ID", "CHR", "BP_START" ,"BP_END", "STRAND", "GENE_NAME" ))
test <- as_tibble(test)

# How many Genes are on the positive strand? 9819
sum(str_count(string = test$STRAND, pattern = "\\+"))

# or with pipes
test %>%
  pull(STRAND) %>%
  stringr::str_count(pattern = "\\+") %>%
  sum()

# How many genes begin and end with a letter?: 5016
sum(str_count(string = test$GENE_NAME, pattern = "^[a-zA-Z].*[a-zA-Z]$"))

# or with pipes again
test %>%
  pull(GENE_NAME) %>%
  stringr::str_count(pattern = "^[a-zA-Z].*[a-zA-Z]$") %>%
  sum()
```