---
title: "MET581 Lecture 05 Homework Answers"
subtitle: "Wrangling Data 3"
author: "Matthew Bracher-Smith"
date: "2024-10-21"
format: 
  html:
    number-sections: true
    theme: united
    highlight: tango
    mainfont: Ubuntu
    embed-resources: true
    self-contained: true
editor: source
execute:
  echo: true
---

```{r}
#| output: false
#| warning: false
library(forcats)
library(dplyr)
library(stringr)
```

This document contains all questions and answers for the lecture 'Wrangling Data 3'. Please create a Quarto document containing all text, code and output used to answer the questions.

## Factors
1. When does `fct_lump()` stop adding levels into “other”?

_It combines as many categories into 'other' as it can while keeping 'other' as the smallest category. Users can also specify limits using `n` and `prop`. fct_lump has largely been superseded by more specific versions such as `fct_lump_min`, `fct_lump_prop` etc., but the original `fct_lump` is still supported and is the most common version you will see._

2. Load the `gss_cat` dataframe and get a summary to see if the amount of tv hours per day affects the political party a person belongs to.

```{r}
#| warning: false
# we can get summaries of tv hours by politcal party without needing forcats
# note you may also use a combination of summarise and across
gss_cat |>
  group_by(partyid) |>
  summarise_at(vars('tvhours'),
               funs(sum(., na.rm = TRUE), mean(., na.rm = TRUE),
                    median(., na.rm = TRUE)))

# we could also consider making tvhours categorical, lumping together
# extreme numbers of hours, and breaking-down the number of parties by hours
gss_cat |>
  select(tvhours, partyid) |>
  na.omit() |>
  mutate(tvhours = fct_lump(as.factor(tvhours), n=7,
                            other_level = '7+')) |>
  group_by(tvhours) |>
  dplyr::count(partyid)
```

_Note that whatever the exploratory breakdown, we cannot actually answer a causal question with this data/method, like whether 'the amount of tv hours per day affects the political party a person belongs to'._

3. Repeat the previous exercise, but this time merge all the “other party” levels into a single option and reorder the factor by the average tv hours per day.

```{r}
# assuming that all "other party" levels means all no democrat/republicans
# it doesn't matter which parties they combined, as long as students could
# combine factor levels using fct_collapse() and reorder using fct_reorder()
other_parties <- c("Other party", "Independent", "Ind,near rep", "Ind,near dem")
new_cat <- gss_cat |>
  mutate(partyid = fct_collapse(partyid,
                                `Other or Independent` = other_parties)) |>
  group_by(partyid) |>
  summarise_at(vars('tvhours'),
               funs(sum(., na.rm = TRUE), mean(., na.rm = TRUE),
                    median(., na.rm = TRUE))) |>
  mutate(partyid = fct_reorder(partyid, mean))

levels(new_cat$partyid)
```

_You have a vector of categorical data that has a natural order to it, and you want to use in a regression model. The vector is c('Some', 'All', 'None', 'Half', 'Most', 'Most', 'Most', 'All', 'Some', 'None'). Create a factor from this vector by pasting it into the `factor()` function and assign to a variable._

4. You're concerned to make sure that 'None' is set as the baseline. Check the coding for regression problems with `contrasts()` and print the output.

```{r}
data <- factor(c('Some', 'All', 'None', 'Half', 'Most', 'Most', 'Most', 'All', 'Some', 'None'))
contrasts(data)
```

5. Why are the levels in this order by default?

_When no levels are passed, levels are created in alphabetical order._

6. Re-order the levels of this factor to go from 'None' to 'All' in increasing amount
```{r}
# students may have re-created the factor afresh
data <- factor(c('Some', 'All', 'None', 'Half', 'Most', 'Most', 'Most', 'All', 'Some', 'None'),
               levels = c('None', 'Some', 'Half', 'Most', 'All'))

# using forcats::fct_reorder()
# the preferred way, as the factor was already created
data <- fct_relevel(data, c('None', 'Some', 'Half', 'Most', 'All'))
```

## Functions
7. Can you tell which of the arguments of myFunction are mandatory and which are optional?
```{r}
#| eval: false
myFunction <- function(x, y, verbose = FALSE){
  result <- x ** y
  if (verbose){
    print(result)
  }
  return(result)
}
```

_**x** and **y** are mandatory, but **verbose** is optional because a default value has already been assigned in the function definition._

8. What will be the output of the previous function if we type the following command myFunction(2, 3, TRUE)? Try to do it without running the code.

_There will be two outputs of 8. One is printed, because verbose is set to TRUE. The other is the return value of the function, which is automatically printed if we do not assign the output of a function to a variable._

9. Write `any_na()`, a function that takes two vectors of the same length and returns the number of positions that have an NA in, at least, one of the vectors.
```{r}
#| eval: false
# multiple options are available for ensuring the function takes arguments that are
# the same length. Extra points if they used a package like `assertthat`

# code below is what students might write to count all the NAs across both vectors
any_na <- function(x, y){
  stopifnot(length(x) == length(y))
  
  return(sum(c(is.na(x), is.na(y))))
}

# however, we want the number of positions, not number of total NAs
# again. There are many possible solutions. A simple one is to make a quick
# dataframe, as they have many functions available for NA operations
any_na <- function(x, y){
  # returns the number of positions that have an NA in at least one of the vectors
  stopifnot(length(x) == length(y))
  
  return(sum(!complete.cases(data.frame(x, y))))
}
```

_If any intermediate variables were created, they should have sensible names. A comment explaining what the function does is also welcome! Because we know nothing about the problem application for this question, something like x and y can be reluctantly accepted as names for input variables._

10. Write `complementary()`, a function that takes a DNA strand of variable length and returns its complementary strand.
```{r}
#| eval: false
# one option is using chartr, a function from base R built for character translation
complementary <- function(dna_seq){
  # returns complimentary strand for variable-length DNA strands
  return(chartr("ATGC","TACG", dna_seq))
}

# it wasn't specified what the strand would be in
# if we supplied the sequence as a vector, vectorised ifelse
# statement may be handy
complementary <- function(dna_seq){
  # returns complimentary strand for variable-length DNA strands
  return(dplyr::case_when(dna_seq == 'A' ~ 'T',
                          dna_seq == 'T' ~ 'A',
                          dna_seq == 'C' ~ 'G',
                          dna_seq == 'G' ~ 'C'))
}
```

_Whatever was used, variable and argument names should be appropriate. X or y are no longer acceptable, as we know something about the problem for this question._

11. Write a function that transcribes and translates a DNA strand into its corresponding protein sequence. You can assume that the length of the input DNA strand is a multiple of 3.

```{r}
#| eval: false
# students may have broken the problem into sub functions
# if this is the case, they should not be nested (they should
# not be defined within another function's body)
transcribe <- function(dna_seq){
  # converts from dna to rna
  rna_seq <- str_replace_all(dna_seq, 'T', 'U')
  
  return(rna_seq)
}

# even better than the option below would be putting the RNA codon table
# into a separate file which is loaded, so we can easily update it
translate <- function(rna_seq){
  # converts from RNA to protein
  # define an RNA codon table
  rna_codons = c("F", "L", "I", "V", "F", "L", "I", "V",
                "L", "L", "I", "V", "L", "L", "M", "V",
                "S", "P", "T", "A", "S", "P", "T", "A",
                "S", "P", "T", "A", "S", "P", "T", "A",
                "Y", "H", "N", "D", "Y", "H", "N", "D",
                "STOP", "Q", "K", "E", "STOP", "Q", "K", "E",
                "C", "R", "S", "G", "C", "R", "S", "G",
                "STOP", "R", "R", "G", "W", "R", "R", "G")
  names(rna_codons) <- c("UUU", "CUU", "AUU", "GUU", "UUC",
                         "CUC", "AUC", "GUC", "UUA", "CUA",
                         "AUA", "GUA", "UUG", "CUG", "AUG",
                         "GUG", "UCU", "CCU", "ACU", "GCU",
                         "UCC", "CCC", "ACC", "GCC", "UCA",
                         "CCA", "ACA", "GCA", "UCG", "CCG",
                         "ACG", "GCG", "UAU", "CAU", "AAU",
                         "GAU", "UAC", "CAC", "AAC", "GAC",
                         "UAA", "CAA", "AAA", "GAA", "UAG",
                         "CAG", "AAG", "GAG", "UGU", "CGU",
                         "AGU", "GGU", "UGC", "CGC", "AGC",
                         "GGC", "UGA", "CGA", "AGA", "GGA",
                         "UGG", "CGG", "AGG", "GGG" )
  
  # optionally check length is multiple of 3
  stopifnot((nchar(rna_seq) %% 3) == 0)
  
  # lookup one-letter amino acid code for each codon
  amino_acids <- ''
  for(i in seq(1, nchar(rna_seq), by=3)){
    codon <- substr(rna_seq, i, i+2)
    amino_acids <- paste(amino_acids, as.character(rna_codon[codon]), sep = '')
  }
  
  return(amino_acids)
}

dna_to_protein <- function(dna_seq) {
  # convert from dna to protein
  rna_seq <- transcribe(dna_seq)
  protein_seq <- translate(rna_seq)
  return(protein_seq)
}
```

_As before, variable names and function names should be appropriate. Comments are probably needed to make parts clear. It makes sense for separate operations to be broken into separate functions like above. This format is much-preferred. If anyone did the whole thing in a single function, some effort should be made to clearly separate-out the different sections._