---
title: "Exploratory Data Analysis in R"
author:
  - name: "Gabriel Mateus Bernardo Harrington"
    orcid: 0000-0001-6075-3619
    email: bernardo-harringtong@cardiff.ac.uk
    affiliations:
      - ref: cardiff
      - ref: dri
    degrees:
      - PhD
affiliations:
  - id: cardiff
    name: Cardiff University
    city: Cardiff
    url: https://www.cardiff.ac.uk
  - id: dri
    name: UK Dementia Research Institute
    url: https://ukdri.ac.uk/
format:
  revealjs:
    logo: /./10_resources/00_images/combined_logos.png
    footer: "Data for Life Sciences 1 - Slides and code available [here](https://h-mateus.github.io/MET581---Computing-for-Bioinformatics-and-Genetic-Epidemiology/)"
    theme: [dark, ../styles.scss, brand]
    brand-mode: dark
    margin: 0.05
    drop:
      engine: webr
      webr:
        packages:
          - tidyverse
          - conflicted
revealjs-plugins:
  - drop
execute:
  echo: true
---

## Exploratory Data Analysis {.center}

```{r}
#| include: false
library(tidyverse)
library(conflicted)
conflicts_prefer(dplyr::filter)

cohort <- read_csv(
  here::here("Simulated_data/generated_uk/bioinformatics_msc_stats_baseline.csv"),
  show_col_types = FALSE
) |>
  select(
    participant_id, age_years, sex_at_birth, imd_quintile, smoking_status,
    self_rated_health, height_cm, bmi_kg_m2, systolic_bp_mmHg,
    crp_mg_L, obesity, type2_diabetes
  )
```

## Todays Aims

- Get to know a dataset before you do anything clever with it
- Choose the right summary for the shape of the data you actually have
- Find the problems **before** they reach your results
- Describe relationships between variables honestly

## Learning Objectives

By the end of this session you should be able to:

- Classify a variable as continuous, binary, categorical or ordered
- Summarise each type appropriately, and justify the choice
- Recognise skew, outliers and missingness, and say what to do about them
- Produce and interpret a cross-tabulation and a correlation

## What EDA actually is

Not a step you tick off. It's the habit of **looking at your data before you
trust it**.

::: {.incremental}
- What is in here, and what type is each thing?
- What is the typical value, and how spread out is it?
- What is broken, missing, or impossible?
- What looks related to what?
:::

. . .

Every statistical test you meet in sessions 9–14 assumes you have already done
this. None of them will tell you if you haven't.

## The rule for this session {.smaller}

> **Look at the data. Then look again. Then compute something.**

. . .

The expensive mistakes in data analysis are almost never "I used the wrong test".

. . .

They are "I didn't notice that column was a character", or "I didn't notice two
rows were duplicated", or — as you'll see in about ten minutes — "I didn't notice
two values were ten times too big".

## Our dataset {.smaller}

A **simulated** cohort of 2,400 older adults in England: demographics,
deprivation, lifestyle, and clinical measurements.

- It is entirely made up. No real patients.
- It is the **same dataset** you'll use in the statistics sessions
- It is deliberately **not clean**, because real data never is

```{r}
dim(cohort)
```

::: {.notes}
The full table has 73 columns; this is a teaching subset of 12 so that output
fits on a slide. Show the full one live if there's time.
:::

## First contact {.smaller}

Before anything else — how big is it, and what is in it?

```{r}
glimpse(cohort)
```

## What `glimpse()` just told you {.smaller}

::: {.incremental}
- `<dbl>` — a number (double). Age, BMI, blood pressure.
- `<chr>` — text. Sex, smoking status, self-rated health.
- 2,400 rows, 12 columns
:::

. . .

Note what is **not** there: nothing is a factor yet, and `obesity` is stored as a
number even though it only takes two values.

. . .

R will happily let you take the mean of a 0/1 column, or sort text
alphabetically when you meant "worst to best". Knowing the type is how you avoid
both.

## Variable types {.smaller}

| Type | Looks like | Example here |
|------|-----------|--------------|
| **Continuous** | any value in a range | `age_years`, `bmi_kg_m2` |
| **Binary** | exactly two outcomes | `obesity`, `type2_diabetes` |
| **Categorical** (nominal) | unordered groups | `sex_at_birth`, `smoking_status` |
| **Ordered** (ordinal) | groups with a rank | `imd_quintile`, `self_rated_health` |

. . .

**Classify first, compute second.** The type decides the summary, the plot, and
later the test.

## Practice! {.smaller}

Without running anything, classify each of these:

- `height_cm`
- `sex_at_birth`
- `self_rated_health`
- `type2_diabetes`
- `imd_quintile`

. . .

Then check yourself with `glimpse()` — and notice where R's *storage* type
disagrees with the *conceptual* type.

# Continuous variables

## Central tendency {.smaller}

Two ways to say "the typical value":

```{r}
cohort |>
  summarise(
    mean   = mean(age_years),
    median = median(age_years)
  )
```

. . .

Nearly identical — which tells you the distribution is roughly symmetric.

When they **disagree**, that disagreement is information.

## Variability {.smaller}

"Typical" is useless without "how spread out".

```{r}
cohort |>
  summarise(
    sd    = sd(age_years),
    IQR   = IQR(age_years),
    range = max(age_years) - min(age_years)
  )
```

::: {.incremental}
- **SD** — average distance from the mean. Uses every value.
- **IQR** — width of the middle 50%. Ignores the tails.
- **Range** — max minus min. Entirely decided by two values.
:::

## Now do it for blood pressure {.smaller}

```{r}
cohort |>
  summarise(
    n      = n(),
    mean   = mean(systolic_bp_mmHg),
    median = median(systolic_bp_mmHg),
    sd     = sd(systolic_bp_mmHg),
    IQR    = IQR(systolic_bp_mmHg),
    min    = min(systolic_bp_mmHg),
    max    = max(systolic_bp_mmHg)
  )
```

. . .

**Something here is wrong. What?**

## Spot it {.smaller}

::: {.incremental}
- Mean 140, median 139 — fine, that is a plausible systolic blood pressure
- IQR 17 — fine
- **SD 40.8** — for a variable whose middle 50% spans 17 mmHg?
- **Max 1620** — no.
:::

. . .

A standard deviation more than twice the IQR is a red flag on its own. You did
not need the `max` to know something was wrong.

. . .

This is why you compute **both** a robust and a non-robust summary.

## Look at it {.smaller}

```{r}
#| fig-height: 3.5
boxplot(cohort$systolic_bp_mmHg,
        horizontal = TRUE,
        xlab = "Systolic BP (mmHg)")
```

. . .

The entire distribution is squashed against the left edge by two points.

## Find it {.smaller}

```{r}
cohort |>
  filter(systolic_bp_mmHg > 250) |>
  select(participant_id, age_years, systolic_bp_mmHg)
```

. . .

`1330` and `1620`. Those are `133.0` and `162.0` with a misplaced decimal point —
a data entry error, and a very common one.

## Fix it {.smaller}

```{r}
cohort_clean <- cohort |>
  mutate(
    systolic_bp_mmHg = if_else(
      systolic_bp_mmHg > 250,
      systolic_bp_mmHg / 10,
      systolic_bp_mmHg
    )
  )

cohort_clean |>
  summarise(mean = mean(systolic_bp_mmHg), sd = sd(systolic_bp_mmHg))
```

. . .

SD falls from **40.8** to **12.6**. The mean barely moves.

::: {.callout-warning appearance="simple"}
Never silently delete a row. Record what you changed and why — that is what makes
a pipeline reproducible, and it is marked.
:::

## Robust vs non-robust {.smaller}

Watch which summaries even noticed:

| | with the error | after the fix |
|---|---|---|
| mean | 140.0 | 138.9 |
| **median** | **139** | **139** |
| sd | 40.8 | 12.6 |
| **IQR** | **17** | **17** |

. . .

The median and IQR did not move at all. They are **robust** — built from ranks,
so extreme values carry no extra weight.

. . .

That is the argument for reporting both. Where they disagree, look.

## What it cost {.smaller}

Two bad cells out of 2,400. Here is the correlation between age and blood
pressure, before and after:

```{r}
c(
  raw     = cor(cohort$age_years,       cohort$systolic_bp_mmHg),
  cleaned = cor(cohort_clean$age_years, cohort_clean$systolic_bp_mmHg)
) |> round(3)
```

. . .

**0.09 to 0.29.** A real, moderate association was hidden entirely by two typos.

. . .

Had you skipped EDA you would have reported "no relationship between age and
blood pressure" — confidently, and wrongly.

## Practice! {.smaller}

Using `cohort`:

- Summarise `bmi_kg_m2` with mean, median, SD and IQR. Do any disagree?
- Find the minimum and maximum. Are both plausible for an adult?
- Produce a boxplot of `crp_mg_L`. What do you notice?

**Extra practice**

- Write a small check that flags any column whose SD is more than twice its IQR

## Skew {.smaller}

Compare two continuous variables:

```{r}
cohort |>
  summarise(
    across(
      c(height_cm, crp_mg_L),
      list(mean = ~ mean(.x, na.rm = TRUE), median = ~ median(.x, na.rm = TRUE))
    )
  ) |>
  round(2)
```

. . .

Height: mean ≈ median. Symmetric.

CRP: mean **above** median. A long right tail is dragging the mean up.

## Seeing skew {.smaller}

```{r}
#| fig-height: 3.5
#| layout-ncol: 2
hist(cohort$height_cm, main = "height_cm", xlab = NULL)
hist(cohort$crp_mg_L,  main = "crp_mg_L",  xlab = NULL)
```

## Which summary do you report? {.smaller}

::: {.incremental}
- **Symmetric** — mean and SD are fine, and use all the data
- **Skewed** — median and IQR describe it far better
- Reporting a mean for a strongly skewed variable is not *wrong*, it is
  **misleading**, which is worse
:::

. . .

You will meet this again in sessions 11–12: the same shape decides whether a
parametric test is appropriate.

## Break {.center}

# Missing data

## Where is it missing? {.smaller}

```{r}
cohort |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))
```

## Two different kinds of missing {.smaller}

::: {.incremental}
- **Sporadic** — a blood sample was not taken. A few percent, scattered.
- **Structural** — the value cannot exist for that person.
:::

. . .

In the full dataset, `age_at_death` is missing for 91.6% of people. Not a data
quality problem — those people did not die.

. . .

Treating those two the same way is a classic and expensive error. Always ask
**why** a value is missing before deciding what to do with it.

## `NA` will not warn you {.smaller}

```{r}
mean(cohort$crp_mg_L)
```

. . .

```{r}
mean(cohort$crp_mg_L, na.rm = TRUE)
```

. . .

The first is arguably the better behaviour — R is telling you something is
missing. `na.rm = TRUE` says *"I know, and I've decided to ignore them"*.

Make sure that is actually a decision you made.

# Categorical and binary variables

## Counting {.smaller}

The whole toolkit for a categorical variable starts here:

```{r}
cohort |> count(smoking_status)
```

. . .

**Four** categories for a three-category variable.

## The levels are not clean {.smaller}

```{r}
cohort |> count(self_rated_health, sort = TRUE)
```

. . .

`Very Good`, `Very good` and `very good` are three spellings of one category.
`count()` found them because it does not guess.

## Why this matters {.smaller}

If you had gone straight to a test, `Very good` (136 people) and `very good`
(1 person) would have been compared as **different groups**.

. . .

::: {.incremental}
- Nothing errors
- The output has the right shape
- Your degrees of freedom are quietly wrong
:::

. . .

Repairing these is a job for `case_when()` and `forcats` — session 5, and the
[factors reference page](../10_resources/02_reference/factors_with_forcats.qmd).

## Binary variables {.smaller}

`obesity` has two possible values, so its summary is just a proportion. Let's
take the mean:

```{r}
mean(cohort$obesity)
```

. . .

`NA` — because it is **text**, not 0/1. `glimpse()` told us that earlier: `<chr>`.

. . .

Note that it did **not** error. You get `NA`, a warning you can easily scroll
past, and a pipeline that carries on.

## And it is not even clean {.smaller}

```{r}
cohort |> count(obesity)
```

. . .

**Four** values for a two-value variable. `Y` and `YES` are one person each.

. . .

If you had filtered on `obesity == "Yes"` you would have silently dropped two
people. No error, no warning.

## Clean, then summarise {.smaller}

```{r}
cohort_clean <- cohort_clean |>
  mutate(
    obese = str_to_sentence(str_trim(obesity)),
    obese = obese %in% c("Y", "Yes")
  )

cohort_clean |> count(obesity, obese)
```

## Now the proportion {.smaller}

```{r}
cohort_clean |>
  summarise(
    n          = n(),
    n_obese    = sum(obese),
    prevalence = mean(obese)
  )
```

. . .

`mean()` of a `TRUE`/`FALSE` column **is** the proportion — `TRUE` counts as 1.

. . .

33.8%. Which you could only compute *after* knowing the type and fixing the
levels. That is the entire argument for this session in one slide.

## Practice! {.smaller}

Using `cohort`:

- How many people are in each `imd_quintile`? Is any level unexpected?
- What proportion have `type2_diabetes`?
- Count `sex_at_birth`. Does anything need cleaning?

**Extra practice**

- Write one pipeline that returns the number of distinct values of every
  character column, and use it to spot which need attention

# Relationships between variables

## Cross-tabulation {.smaller}

Two categorical variables at once:

```{r}
cohort_clean |>
  count(sex_at_birth, obese) |>
  pivot_wider(names_from = obese, values_from = n, names_prefix = "obese_")
```

. . .

Counts alone are hard to compare when the groups are different sizes.

## Proportions within groups {.smaller}

```{r}
cohort_clean |>
  group_by(sex_at_birth) |>
  summarise(n = n(), obesity_prevalence = mean(obese)) |>
  arrange(desc(n))
```

. . .

Now the comparison is fair — and look at the third row. The three `female`
records are being reported as a separate group with 0% prevalence.

. . .

One uncleaned capital letter has invented a third sex category with an
eye-catching and completely meaningless result.

::: {.callout-note appearance="simple"}
**Always ask which way the percentages run.** "60% of obese people are female"
and "60% of females are obese" are completely different claims.
:::

## Correlation {.smaller}

For two continuous variables:

```{r}
cor(cohort_clean$age_years, cohort_clean$systolic_bp_mmHg)
```

. . .

Pearson's *r* runs from −1 to +1:

::: {.incremental}
- **sign** — direction. Positive: both rise together.
- **magnitude** — strength. 0 is no *linear* relationship.
:::

. . .

0.29 is a moderate positive association. Older people tend to have higher blood
pressure — but the relationship is loose.

## Correlation with missing values {.smaller}

```{r}
cor(cohort$bmi_kg_m2, cohort$crp_mg_L)
```

. . .

`NA`, because some CRP values are missing.

```{r}
cor(cohort$bmi_kg_m2, cohort$crp_mg_L, use = "complete.obs")
```

. . .

`use = "complete.obs"` drops incomplete pairs. As ever — that is a decision, so
know that you made it.

## What correlation does not tell you {.smaller}

::: {.incremental}
- It does not imply **causation**
- It only measures a **straight-line** relationship
- It is badly distorted by **outliers** — as you saw, 0.09 vs 0.29
- A **significant** correlation can still be a tiny one
:::

. . .

On 2,400 people, *r* = 0.09 is statistically significant. It is also almost
completely useless for predicting anything.

. . .

**Always plot it before you believe it.**

## Plot it {.smaller}

```{r}
#| fig-height: 3.5
plot(cohort_clean$age_years, cohort_clean$systolic_bp_mmHg,
     pch = 16, col = rgb(0.56, 0.66, 1, 0.3),
     xlab = "Age (years)", ylab = "Systolic BP (mmHg)")
```

. . .

A real trend, and a lot of scatter. That is what *r* = 0.29 looks like.

## Practice! {.smaller}

Using `cohort_clean`:

- Correlate `bmi_kg_m2` with `systolic_bp_mmHg`. Plot it. Does the number match
  what you see?
- Compare mean `bmi_kg_m2` across `smoking_status` groups
- Cross-tabulate `imd_quintile` against `type2_diabetes` as proportions

**Extra practice**

- Find the pair of continuous variables in the full dataset with the strongest
  correlation. Is it interesting, or just two versions of the same measurement?

## An EDA checklist {.smaller}

Before any test, on any dataset:

::: {.incremental}
1. `dim()` and `glimpse()` — size and types
2. Classify each variable you plan to use
3. Summarise continuous variables **both** ways — mean/SD and median/IQR
4. `count()` every categorical variable, and read the levels
5. Count the missing values, and ask *why* they are missing
6. Plot anything you intend to claim a relationship about
:::

. . .

It takes ten minutes and it is the difference between an analysis you can defend
and one you cannot.

## Additional Resources

- [R for Data Science 2e — Exploratory Data Analysis](https://r4ds.hadley.nz/eda)
- [`skimr`](https://docs.ropensci.org/skimr/) — fast, readable summaries of a
  whole data frame
- [`naniar`](https://naniar.njtierney.com/) — tools for exploring missing data
- Next session: doing all of this **properly**, with `ggplot2`
