---
title: "Working with AI assistants"
subtitle: "Fluent, confident, and wrong"
author:
  - name: "Gabriel Mateus Bernardo Harrington"
    orcid: 0000-0001-6075-3619
    email: bernardo-harringtong@cardiff.ac.uk
    affiliations:
      - ref: cardiff
      - ref: dri
    degrees:
      - PhD
affiliations:
  - id: cardiff
    name: Cardiff University
    city: Cardiff
    url: https://www.cardiff.ac.uk
  - id: dri
    name: UK Dementia Research Institute
    url: https://ukdri.ac.uk/
format:
  revealjs:
    logo: /10_resources/00_images/combined_logos.png
    footer: "Data for Life Sciences 1 - Slides and code available [here](https://h-mateus.github.io/MET581---Computing-for-Bioinformatics-and-Genetic-Epidemiology/)"
    slide-number: true
    theme: [dark, ../styles.scss, brand]
    brand-mode: dark
    margin: 0.05
    drop:
      engine: webr
      webr:
        packages:
          - tidyverse
revealjs-plugins:
  - drop
execute:
  echo: true
---

## Todays Aims

- Be honest about the fact that you will all use AI assistants
- Show you what they get wrong, and why you can't see it
- Give you a way to use them that leaves you *better* at R, not worse
- Set out what's fine and what isn't in assessed work

::: {.notes}
~15-20 minutes, placed after the first coding exercises in session 3 so they
have enough R to read the examples. Run the code live rather than reading the
output off the slide — the reveal is the whole point.
:::

## Let's start with the awkward bit

I am not going to ban you from using LLMs.

- It would not work
- It would be daft — you'll use them for the rest of your career
- I use them

But I am going to try to convince you that **how** you use them over the next
five weeks decides whether this module was worth your time.

## The employability argument {.smaller}

Suppose the whole of your R skillset is "I can ask a model for R code".

. . .

Your employer can do that. Their intern can do that. The person who didn't do
the MSc can do that.

. . .

**The market rate for a skill everyone has is zero.**

. . .

What is actually scarce — and what people pay for:

- Looking at generated code and knowing it's wrong
- Knowing *which* question to ask in the first place
- Being able to defend an analysis to someone who wants to publish it

That's the job. The typing was never the job.

## So let's test it

Here is R code of the sort an assistant will happily produce.

Some of it is wrong.

**None of it errors.**

::: {.notes}
Emphasise: no error message. That's the danger. Broken code that crashes is a
minor inconvenience. Broken code that returns a plausible number ends up in a
paper.
:::

## Exhibit A {.smaller}

> *"How do I get the mean of a column that has missing values?"*

```{r}
x <- c(1, 2, NA, 4)

mean(x, na.omit = TRUE)
```

. . .

`NA`.

There is no `na.omit` argument to `mean()`. It was swallowed by `...` and
silently ignored. The argument is `na.rm`:

```{r}
mean(x, na.rm = TRUE)
```

. . .

**Why you'd miss it:** `na.omit` is a real R function. It reads perfectly.

## Exhibit B {.smaller}

> *"Convert this column to numbers."*

```{r}
genotype_count <- factor(c("10", "20", "5"))

as.numeric(genotype_count)
```

. . .

`1 2 3`. Those are the *factor level codes*, not your data.

```{r}
as.numeric(as.character(genotype_count))
```

. . .

**Why you'd miss it:** you asked for numbers and you got numbers. Small
integers, in a plausible range. No warning. This one has made it into published
papers.

## Exhibit C {.smaller}

> *"Filter out the samples from group a."*

```{r}
#| message: false
library(tidyverse)

df <- tibble(group = c("a", "b", NA, "a"), value = 1:4)

df |> filter(group != "a")
```

. . .

One row. But **two** rows aren't group `"a"` — the `NA` vanished too.

`NA != "a"` isn't `TRUE`, it's `NA`, and `filter()` keeps only `TRUE`.

```{r}
df |> filter(group != "a" | is.na(group))
```

## Exhibit D — the expensive one {.smaller}

> *"Join the phenotype table onto the sample table."*

```{r}
samples <- tibble(id = c(1, 2, 3), sample = c("s1", "s2", "s3"))
pheno   <- tibble(id = c(1, 1, 2), measure = c(4.1, 9.9, 5.2))

joined <- left_join(samples, pheno, by = "id")
```

. . .

```{r}
nrow(samples)
nrow(joined)
```

. . .

3 rows in, **4 rows out**. `id == 1` appears twice in `pheno`, so sample `s1` got
duplicated. Every downstream mean is now wrong, and weighted towards whichever
samples happened to have duplicate records.

## Exhibit D, continued

```{r}
joined
```

. . .

**The habit that saves you:** check `nrow()` before and after every join. Every
single one.

An assistant will almost never do this unprompted. You have to.

## Exhibit E {.smaller}

> *"Apply my function across the list."*

```{r}
above_two <- function(x) x[x > 2]

list_a <- list(a = c(1, 3, 4), b = c(1, 5, 6))
class(sapply(list_a, above_two))
```

Works — a matrix, as you'd hope. Now change one value:

```{r}
list_b <- list(a = c(1, 3, 4), b = c(1, 1, 6))
class(sapply(list_b, above_two))
```

. . .

A **list**. `sapply()` guesses its return type from the data, so your pipeline
breaks on a dataset you haven't seen yet — in three weeks, in a different script.

## Exhibit E — the fix {.smaller}

Type-stable alternatives fail *loudly and immediately* instead:

```{r}
#| error: true
vapply(list_b, above_two, numeric(2))
```

. . .

This is a general principle worth more than the specific example:

> **Prefer the tool that breaks now over the tool that breaks quietly later.**

`purrr::map_dbl()`, `vapply()`, `if_else()` over `ifelse()` — all the same idea.

## Exhibit F {.smaller}

> *"Flag the dates after March."*

```{r}
dates <- as.Date(c("2026-01-01", "2026-06-01"))

ifelse(dates > as.Date("2026-03-01"), dates, NA)
```

. . .

`20605`. Your dates are now integers — `ifelse()` strips the class.

```{r}
if_else(dates > as.Date("2026-03-01"), dates, as.Date(NA))
```

## Exhibit G {.smaller}

> *"Add a column with the mean of a and b."*

```{r}
d <- tibble(a = c(1, 2), b = c(3, 4))

d |> mutate(m = mean(c(a, b)))
```

. . .

Every row got `2.5` — the mean of *all four numbers*. You wanted rowwise.

```{r}
d |> mutate(m = (a + b) / 2)
```

. . .

**Why you'd miss it:** the code says `mean`, you wanted a mean, and a number
appeared. With 10,000 rows you would never look.

## What all seven have in common {.smaller}

::: {.incremental}
- No error. No warning.
- Output of the right *shape* and a plausible *magnitude*
- The code reads like English and looks like what you asked for
- Every one of them is invisible unless you already understand R
:::

. . .

The model isn't lying to you. It's producing the most plausible-looking code, and
plausible-looking code is exactly the thing you cannot audit by looking at it.

## Which is the actual argument {.smaller}

You cannot check what a model gives you unless you know:

- what a factor is, and what's underneath one
- what `NA` does to a comparison
- what a join does to row counts
- what "type stable" means and why you'd want it

**That's this module.** Not because R is precious, but because that knowledge is
the only thing standing between you and a confidently wrong result.

. . .

The people who get burned by these tools aren't the ones who use them. They're
the ones who can't check them.

## There is a second reason {.smaller}

Everything so far was about the model being **wrong**.

. . .

This one applies even when it is **right**.

::: {.incremental}
- Reading a correct solution feels almost exactly like working one out
- That feeling is not learning. It is *fluency*, and it is gone by Thursday
- The struggle is not an obstacle to the learning. It **is** the learning
:::

. . .

Learning scientists call this a *desirable difficulty*: the effort of retrieving
something yourself is what makes it stick. Watching it appear costs you that,
and costs you it silently.

## The loop {.smaller}

::: {.incremental}
1. You get stuck, and you ask
2. You get a working answer, and you move on
3. You never build the knowledge that answer was made of
4. Next time you are **less** able to tell a good answer from a bad one
5. So you ask sooner
:::

. . .

Nothing in that loop feels bad while it is happening. That is precisely the
problem with it.

. . .

And you still sit the same paper in week 5, on your own.

## Three tools, three risk profiles {.smaller}

You have access to **Gemini**, **NotebookLM** and **Elicit**. They are not
interchangeable. They differ in *how much room they have to invent*.

| Tool | Grounded in | Best for |
|------|-------------|----------|
| NotebookLM | sources **you** upload | revising from your own notes |
| Elicit | real published papers | finding literature for the report |
| Gemini | nothing in particular | general help, with every caveat so far |

. . .

Narrower grounding means less room to invent. It does **not** mean "safe".

## The catch with each {.smaller}

**NotebookLM** — grounding cuts invention, it does not remove it. It can still
misread the source you handed it. Good for *"what did we say about joins?"*.
No use at all for *"why will my code not run?"*

. . .

**Elicit** — a 2026 feasibility study in *Research Synthesis Methods* re-ran the
same extractions from different accounts and compared them:

::: {.incremental}
- **90%** agreement on the extracted *values*
- **46%** on the supporting *quotes*
- **30%** on the *reasoning*
:::

. . .

An answer you cannot reproduce is not evidence — and this module is about
reproducible pipelines. Use it to *find* papers. Then read them.

## Break {.center}

## Right — so use them well {.smaller}

Three things that make a genuine difference.

::: {.incremental}
1. **Give it your data, not your description of your data**
2. **Constrain the answer**
3. **Ask it to argue with itself**
:::

## 1. Give it your data {.smaller}

Most bad generated R comes from the model guessing your data's shape. So don't
make it guess:

```{r}
#| eval: false
glimpse(df)
str(df)
dput(head(df, 5))   # paste this - it's a reproducible copy of your data
```

. . .

Better still, tell it the things `str()` can't show:

> *"A row is one participant-visit, so IDs repeat. `measure` is in mg/dL.
> `-99` means missing in the `age` column."*

That single sentence prevents Exhibit D.

## 1. Give it your data — R-native tooling {.smaller}

You can skip the copy-paste entirely. The
[`btw`](https://posit-dev.github.io/btw/) package hands your live R session to
an assistant — actual data frames, actual package versions, actual docs:

```{r}
#| eval: false
install.packages("btw")
library(btw)
btw()          # describes your session for pasting
```

Also worth knowing about: [`ellmer`](https://ellmer.tidyverse.org/) for calling
models from R, and the assistant built into
[Positron](https://positron.posit.co/).

. . .

An assistant that can see your session is dramatically less likely to invent
a column that doesn't exist.

## 2. Constrain the answer {.smaller}

Vague question, average answer. Compare:

::: {.callout-note appearance="simple"}
## Weak
"How do I summarise this data?"
:::

::: {.callout-tip appearance="simple"}
## Better
"Using dplyr only — no new packages — give me mean and SD of `measure` by
`treatment_group`, keeping groups with fewer than 5 observations visible rather
than dropped. Tell me what happens to `NA`s in `measure`."
:::

. . .

Notice what the second one is doing: it names the tools, the output, **and the
edge case you're worried about**. Precision about edge cases is the whole skill.

## 3. Ask it to argue with itself {.smaller}

The highest-value prompts you can learn:

::: {.incremental}
- *"What assumptions did you make about my data?"*
- *"How would this fail? Give me three ways."*
- *"What would you check to prove this worked?"*
- *"Is there a tidyverse function that already does this?"*
- *"You used `sapply` — what happens if the results have different lengths?"*
:::

. . .

Models are far better at critiquing code than at writing it correctly first
time. Use that asymmetry.

## Tutor mode {.smaller}

There's a prompt on the module site that turns your assistant into something
closer to a lab demonstrator: asks what you tried, hints instead of solving,
makes you predict the output before it shows you.

**Module site → Working with AI assistants**

Paste it into Claude's or ChatGPT's custom instructions. There's a Claude Code
version too.

. . .

You can defeat it in four seconds by opening another tab. I know. It's not a
fence — it's there for the days you'd rather learn the thing than have it done
for you.

## What's fine, and what isn't {.smaller}

**Fine — genuinely, use the tools:**

- Explaining errors you can't parse
- "What does this function do?"
- Debugging code *you wrote*
- Sanity-checking your reading of a result
- Tightening your writing
- Learning what's out there

## What's fine, and what isn't {.smaller}

**Not fine in assessed work:**

- Generating the analysis code you submit
- Having it pick your statistical approach
- Having it write your interpretation

. . .

Not because of purity. Because **the 50% report is marked on defending your
analytical choices** (MLO-5), and you cannot defend a choice you didn't make.
That conversation goes badly in a way that is very obvious from the outside.

## The rule of thumb

::: {.callout-important appearance="simple"}
## Attempt → ask → diff
Try it yourself first. Then ask. Then compare what you wrote to what it wrote,
and work out *why* they differ.
:::

That third step is where the learning actually lives, and it's the one everyone
skips.

## One last thing {.smaller}

Nobody gets stronger watching a forklift.

. . .

Five weeks is not long. The version of you that finishes this module having
struggled through the wrangling exercises is materially more employable than the
version that had them generated — not because of virtue, but because only one of
those two can look at Exhibit D and spot it.

. . .

I would rather you produced worse code this month and were better at R in
October.

## Additional Resources

- Module site → **Working with AI assistants** (tutor prompt, prompting guide)
- [`btw`](https://posit-dev.github.io/btw/) — give an assistant your R session
- [`ellmer`](https://ellmer.tidyverse.org/) — call LLMs from R
- [R for Data Science (2e)](https://r4ds.hadley.nz/) — still the best answer to
  most R questions
- `?function_name` — faster than asking, and always correct
