---
title: "Working with AI assistants"
subtitle: "How to use LLMs on this module without wasting the module"
author: "Gabriel Mateus Bernardo Harrington"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: false
execute:
  echo: true
---

## The short version

You may use AI assistants on this module. There is no ban, and I'm not
interested in policing it.

But: **the market rate for "can ask a model for R code" is zero**, because
everyone can do that. Your value — in the rest of this MSc, in your project, and
in a job — is in being able to look at generated code and know whether it's
right. That capability is built by writing R, not by reading R.

So the deal is:

::: {.callout-important appearance="simple"}
## Attempt → ask → diff
1. **Attempt** it yourself first.
2. **Ask** if you're stuck.
3. **Diff** — compare what you wrote to what you were given, and work out *why*
   they differ.
:::

Step 3 is where learning happens and it's the one everybody skips.

## Why this matters more than it sounds

Here is R code of the kind an assistant will cheerfully produce. It does not
error. It returns numbers of the right shape and a plausible size. It is wrong.

```{r}
#| message: false
library(tidyverse)
```

### The argument that doesn't exist

```{r}
x <- c(1, 2, NA, 4)
mean(x, na.omit = TRUE)
```

`na.omit` is not an argument to `mean()`. It was absorbed by `...` and ignored.
The argument is `na.rm`. No warning was issued.

### The factor that became its own level codes

```{r}
counts <- factor(c("10", "20", "5"))
as.numeric(counts)
```

`1 2 3` — the internal level codes, not your data. You need
`as.numeric(as.character(counts))`. This class of bug has reached publication
more than once.

### The join that quietly duplicated your samples

```{r}
samples <- tibble(id = c(1, 2, 3), sample = c("s1", "s2", "s3"))
pheno   <- tibble(id = c(1, 1, 2), measure = c(4.1, 9.9, 5.2))

joined <- left_join(samples, pheno, by = "id")
c(before = nrow(samples), after = nrow(joined))
```

Three rows in, four out, because `id == 1` was duplicated in `pheno`. Every mean
you compute downstream is now wrong. **Check `nrow()` before and after every
join** — an assistant will rarely do this for you.

### The filter that silently ate your missing data

```{r}
df <- tibble(group = c("a", "b", NA, "a"), value = 1:4)
df |> filter(group != "a")
```

One row, not two — `NA != "a"` evaluates to `NA`, and `filter()` keeps only
`TRUE`. If you wanted the missing group kept:
`filter(group != "a" | is.na(group))`.

::: {.callout-note}
There are more of these in the session 3 slides. The common thread is that
**none of them error**. Code that crashes is a minor annoyance; code that
returns a confident wrong number is how bad science happens — and you cannot
spot any of them without understanding factors, `NA` semantics, and joins.

That understanding is what this module is for.
:::

## Tutor mode

The most useful thing you can do is stop your assistant from answering so fast.

There's a paste-in prompt in
[`r_tutor_prompt.md`](r_tutor_prompt.md) that reconfigures any assistant into
something closer to a good lab demonstrator: it asks what you've tried, gives
you the smallest useful hint rather than the solution, makes you predict output
before showing it, and caps how much code it writes unasked. It has an escape
hatch for when you're genuinely out of time.

Tool-specific versions:

| Tool | File | Where it goes |
|---|---|---|
| Any chat assistant | `r_tutor_prompt.md` | Custom instructions / project instructions / first message |
| Claude Code | `skills/r-tutor/SKILL.md` | `~/.claude/skills/r-tutor/SKILL.md` |
| A project folder | `CLAUDE.md.example` | Save as `CLAUDE.md` in your coursework folder |

You can defeat all of this by opening a different tab, and I know that. It isn't
a fence. It's there for the days when you'd rather learn the thing.

## Prompting for R, specifically

### 1. Give it your data, not your description of your data

Most bad generated R comes from the model guessing at your data's shape. Don't
make it guess:

```{r}
#| eval: false
glimpse(df)
str(df)
dput(head(df, 5))   # a reproducible copy of your data - paste this
```

Then add what `str()` can't show:

> "A row is one participant-visit, so IDs repeat. `measure` is mg/dL. `-99`
> means missing in `age`."

That one sentence prevents the duplicated-join bug above.

Better still, let R do it. The [`btw`](https://posit-dev.github.io/btw/) package
hands your live session — real data frames, real package versions, real help
pages — to an assistant:

```{r}
#| eval: false
install.packages("btw")
library(btw)
btw()
```

Also worth your time: [`ellmer`](https://ellmer.tidyverse.org/) for calling
models from R, and the assistant built into
[Positron](https://positron.posit.co/).

### 2. Constrain the answer

Vague question, average answer.

::: {.callout-note appearance="simple"}
## Weak
"How do I summarise this data?"
:::

::: {.callout-tip appearance="simple"}
## Better
"Using dplyr only, no new packages: mean and SD of `measure` by
`treatment_group`. Keep groups with fewer than 5 observations visible rather
than dropping them. Tell me what happens to `NA`s in `measure`."
:::

The second names the tools, the output *and the edge case you're worried about*.
Being precise about edge cases is most of the skill.

### 3. Ask it to argue with itself

Models are much better at criticising code than at writing it correctly first
time. Exploit that:

- "What assumptions did you make about my data?"
- "How could this fail? Give me three ways."
- "What would you check to prove this worked?"
- "Is there a tidyverse function that already does this?"
- "You used `sapply()` — what happens if the results have different lengths?"

### 4. Trust the docs over the model

If an assistant tells you an argument exists, `?function_name` will tell you
whether it does. That check takes three seconds and models hallucinate
tidyverse arguments constantly.

### 5. Notice what it doesn't do

Assistants almost never volunteer: row counts around joins, `NA` handling,
checking whether a column is really the type you think, plotting the data before
modelling it. Those omissions are yours to catch.

## Optional: wiring an assistant into your R session

::: {.callout-warning appearance="simple"}
## Leave this one until later
Everything in this section is **optional and unsupported**. I can't debug it for
you in a lab session, and on a managed university laptop it may not be permitted
at all. Come back to it once you can write R unaided. It's documented here
because it exists and you'd find it anyway — not because you need it to pass.
:::

Look again at the four broken examples near the top of this page. They fail for
one shared reason: **the model was guessing at your data.** It had no way to know
that `pheno` contained a duplicated `id`, that `group` held an `NA`, or that
`counts` was a factor rather than a number.

You can't prompt your way out of that. The structural fix is to stop making it
guess.

### Level 1 --- `btw`

The lower-friction option, and the one to try first.
[`btw`](https://posit-dev.github.io/btw/) collects context from your live R
session — the actual structure of your data frames, your installed package
versions, help pages — and puts it on the clipboard for you to paste into any
assistant.

```{r}
#| eval: false
install.packages("btw")
library(btw)

btw()          # describes your session and environment
btw(pheno)     # describes one specific object
```

No configuration, no editing JSON files, and it works with a plain chat window.
For most people this is the whole benefit at a fraction of the effort.

### Level 2 --- `mcptools`

[`mcptools`](https://posit-dev.github.io/mcptools/) goes further: it lets an
assistant *run code* in your live R session, so it can inspect objects and read
help pages itself rather than being told about them.

```{r}
#| eval: false
install.packages("mcptools")
```

Then register your R session so an assistant can find it. Adding it to your
`.Rprofile` means every session registers automatically:

```{r}
#| eval: false
usethis::edit_r_profile()   # then add the line below to that file
mcptools::mcp_session()
```

Finally, point your assistant at R. For Claude Code, one command:

```bash
claude mcp add -s "user" r-mcptools -- Rscript -e "mcptools::mcp_server()"
```

For Claude Desktop and most other MCP clients, add this to the client's config
file instead:

```json
{
  "mcpServers": {
    "r-mcptools": {
      "command": "Rscript",
      "args": ["-e", "mcptools::mcp_server()"]
    }
  }
}
```

Restart both R and the assistant afterwards.

### The catch, which is real

Two things to be honest about.

**This makes it much easier to skip step 1.** An assistant that can see your data
and run your code is a very short path from "I have a problem" to "I have an
answer", and *attempt → ask → diff* only works if you actually attempt. Pair this
with [tutor mode](#tutor-mode) or you have simply built yourself a faster way to
not learn R.

**A better-informed assistant is more convincingly wrong.** Session access
removes one whole category of error — the invented column, the misjudged data
shape. It removes none of the others. It will not tell you that a t-test was the
wrong choice, that your groups aren't independent, or that the effect you found
is an artefact of how you filtered. Those are the judgements the report is marked
on, and they are still entirely yours.

## What's fine and what isn't

**Fine — genuinely, use the tools:**

- Explaining an error you can't parse
- "What does this function do?"
- Debugging code *you* wrote
- Sanity-checking your interpretation of a result
- Improving your writing
- Finding out what packages exist for a problem

**Not fine in assessed work:**

- Generating the analysis code you submit
- Having it choose your statistical approach
- Having it write your interpretation

This isn't about purity. The 2000-word report is explicitly marked on
**defending your analytical choices and methodological approach** (MLO-5). You
cannot defend a choice you didn't make, and the attempt is transparent from the
outside.

If you're unsure whether something crosses a line, ask me. Asking is never the
wrong move.

## Declaring what you used

For the summative report, you'll be asked for a short **process appendix** — a
few hundred words, unmarked for style:

- what you used an assistant for
- one thing it got wrong or that you rejected, and how you spotted it
- what you changed and why

This is not a confession. Honest, specific answers here demonstrate exactly the
evaluative skill the module is teaching, and "I used it to debug my
`pivot_longer()` call and it suggested a column that didn't exist" is a better
answer than "I didn't use it".

## Further reading

- [R for Data Science (2e)](https://r4ds.hadley.nz/) — still the best answer to
  most R questions
- [`btw`](https://posit-dev.github.io/btw/) — give an assistant your R session
- [`ellmer`](https://ellmer.tidyverse.org/) — LLMs from R
- [`mcptools`](https://posit-dev.github.io/mcptools/) — let an assistant run code
  in your live R session (optional, see above)
- `?function_name` — faster than asking, and always correct
